Creating spectroscopic templates using heterogeneous data

### 2005-05-26 docs/paper/fitting.tex
add
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@387 75dff457-8751-0410-ab1a-cd701654ce6b
add
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@402 75dff457-8751-0410-ab1a-cd701654ce6b

\documentclass[10pt]{aastex}

@@

\title{ Using heterogeneous data to constrain star-formation histories
of galaxies}

@@

\begin{abstract}
Astronomers have a huge set of well-calibrated photometric and
spectroscopic data on galaxies from recent large surveys at all
redshifts, such as (at low redshift) the GALaxy Evolution eXplorer in
the ultraviolet, the Sloan Digital Sky Survey in the optical, and the
Two-Micron All Sky Survey in the near infrared, and (at higher
redshifts) the Deep Extragalactic Evolutionary Probe 2 and the Great
Observatories Origins Deep Survey.  However, this data comes in a
rather heterogeneous form --- sometimes in the form of spectra,
sometimes in the form of broad-band photometry. In addition,
broad-band photometry suffers from two forms of heterogeneity: system
responses differ among different surveys, and galaxies are observed at
many different redshifts (and the system responses therefore differ
among all galaxies in the rest frame).  How can we explore this very
heterogeneous set of data?  Here we present an implementation of an
algorithm for fitting star-formation histories to these data. Our
approach is to restrict the problem by trying to explain the
observations of each galaxy as a nonnegative linear combination of a
small set of star-formation history templates.  The algorithm has
almost arbitrary freedom in choosing the set of templates, but it must
explain every galaxy in terms of a nonnegative sum of the same set.
Our algorithm can handle whatever spectral or broad-band information
one has on each galaxy.  The results of the algorithm have scientific
applications --- such as exploring the star-formation histories of
galaxies, how they vary with environment, how they correlate with
other properties, and how they change with redshift. In addition, they
have technical applications --- such as $K$-corrections, photometric
redshifts, and continuum-fitting of spectra. This algorithm and its
ilk are likely to become more and more useful as astronomers continue
to combine large sets of inhomogeneous data.
\end{abstract}

@@

On the other hand, from the spectroscopic observations we know the
galaxy spectra reside in a low-dimensional locus. Principal Component
Analysis (PCA) of galaxy spectra, introduced by \citet{connolly94a} and
applied many times since then ({\bf more refs}; \citealt{yip05a}),
demonstrates that most of the variance in the distribution of galaxies
in spectral space can be explained using a few templates. This means,
that in the linear space of all possible spectra, galaxies exist in
only a small local. Therefore, even with very heterogeneous data, we
should be able to determine the properties of this locus.

Here we present an approach to combining the heterogeneous data
described above in order to determine the properties of the locus of
galaxy spectra. Rather than taking the model-free approach used by
PCA, we here restrict the space of possible spectra to those predicted
from the high resolution stellar population synthesis model of
\citet{bruzual03a}. This approach both constrains the problem
appropriately and yields a natural theoretical interpretation of the
results in terms of star-formation histories.

@@

This approach is similar to PCA in that it finds the small spectral
subspace in which galaxies exist, and can in some ways be thought of
as ``nonnegative PCA.''  However, our method has several advantages
over standard PCA approach. First, it has a natural physical
interpretation associated with that spectral subspace, which is the
corresponding subspace of all possible star-formation
histories. Second, it naturally handles data uncertainties and missing
data which allows it to ignore variation which is due purely to
statistical errors.  Third, it handles the complications of observing
galaxy spectra photometrically using broad-band filters.

@@

Our model for the star-formation history of each galaxy $i$ is a linear
combination of $N$ templates, as follows:
\begin{equation}
G_i(t, m, d) = \sum_{i=0}^{N-1} a_{ik} T_k(t, m, \tau_V).
\end{equation}
We restrict to nonnegative star-formation rates ($a_{ik} \ge
0$). $T_k(t, m, \tau_V)$ is the star-formation history, described as
the stellar mass (in solar masses) formed $t$ years ago, with
metallicity $m$, and currently obscured by $\tau_V$ magnitudes in the
$V$ band.

Each template is in turn expressed as a linear combination of a set of
instantaneous bursts:
\begin{equation}
T_k(t, m, d) = 
\sum_{l=0}^{N_t-1} 
\sum_{n=0}^{N_m-1} 
\sum_{p=0}^{N_\tau-1} 
t_{lmn} 
\delta^D(t-t_l)
\delta^D(m-m_n)
\delta^D(\tau_V-\tau_{V,p})
\end{equation}

\subsection{Predicted spectra from the star-formation histories}
\label{spectra}

Note that we do not {\it have} to limit our spectral library to those
predicted by star-formation histories. In fact, we can add arbitrary
spectra in this step to account for emission lines, polycyclic
aromatic hydrocarbon (PAH) emission, active galactic nuclei, or any
other constituent of galaxies.

In fact, as an example of this, we here include nine emission lines:
OII 3728 \AA; H$\beta$ 4863 \AA; OIII 4960, 5008 \AA; NII 6550, 6585
\AA; H$\alpha$ 6565 \AA; SII 6718, 6733 (all vacuum wavelengths). We
include each of these lines as a template with a Gaussian centered on
the line center with a dispersion corresponding to 300 km s^{-1}.

@@

\subsection{Finding the best-fit templates}
\label{nmf}

@@

%\include{tables}

\include{figures}

>>>>

\documentclass[10pt,preprint]{aastex}

@@

\title{ Creating spectroscopic templates using heterogeneous data}

@@

\begin{abstract}
Astronomers have a huge set of well-calibrated photometric and
spectroscopic data on galaxies from recent large surveys at all
redshifts, such as (at low redshift) the GALaxy Evolution eXplorer in
the ultraviolet, the Sloan Digital Sky Survey in the optical, and the
Two-Micron All Sky Survey in the near infrared, and (at higher
redshifts) the Deep Extragalactic Evolutionary Probe 2 and the Great
Observatories Origins Deep Survey.  However, this data comes in a
rather heterogeneous form --- sometimes in the form of
spectrophotometry, sometimes in the form of broad-band photometry. In
addition, broad-band photometry suffers from two forms of
heterogeneity: system responses differ among different surveys, and
galaxies are observed at many different redshifts (and the system
responses therefore differ among all galaxies in the rest frame).  How
can we explore this very heterogeneous set of data?  Here we present
an implementation of an algorithm for fitting star-formation histories
to these data. Our approach is to restrict the problem by trying to
explain the observations of each galaxy as a nonnegative linear
combination of a small set of star-formation history templates.  The
algorithm has almost arbitrary freedom in choosing the set of
templates, but it must explain every galaxy in terms of a nonnegative
sum of the same small set.  Our algorithm can handle whatever spectral
or broad-band information one has on each galaxy.  The results of the
algorithm have scientific applications --- such as exploring the
star-formation histories of galaxies, how they vary with environment,
how they correlate with other properties, and how they change with
redshift. In addition, they have technical applications --- such as
$K$-corrections, photometric redshifts, and continuum-fitting of
spectra. This algorithm and its ilk are likely to become more and more
useful as astronomers continue to combine large sets of inhomogeneous
data.
\end{abstract}

@@

On the other hand, from the spectroscopic observations we know the
galaxy spectra reside in a low-dimensional locus. Principal Component
Analysis (PCA) of galaxy spectra, introduced by \citet{connolly94a}
and applied many times since then ({\bf more refs}; \citealt{yip05a}),
demonstrates that most of the variance in the distribution of galaxies
in spectral space can be explained using a few templates. This means
that, in the linear space of all possible spectra, galaxies exist in
only a small locus. Therefore, even with very heterogeneous data, we
should be able to determine the properties of this locus.

Here we present an approach to combining the heterogeneous data
described above in order to determine the properties of the locus of
galaxy spectra. Rather than taking the model-free approach used by
PCA, we here restrict the space of possible spectra to those predicted
from the high resolution stellar population synthesis model of
\citet{bruzual03a} and the emission line models of {\bf mappings
ref}.  This approach both constrains the problem appropriately and
yields a natural theoretical interpretation of the results in terms of
star-formation histories.

@@

This approach is similar to PCA in that it finds the small spectral
subspace in which galaxies exist, and can in some ways be thought of
as ``nonnegative PCA.''  However, our method has several advantages
over the standard PCA approach. First, it has a natural physical
interpretation associated with that spectral subspace, which is the
corresponding subspace of all possible star-formation
histories. Second, it naturally handles data uncertainties and missing
data, which allows it to ignore that variation which is due purely to
statistical errors.  Third, it handles the complications of observing
galaxy spectra photometrically using broad-band filters.

@@

Our model for the star-formation history of each galaxy $i$ is a linear
combination of $N$ templates, as follows:
\begin{equation}
\label{stellar}
G_i(t, m, d) = \sum_{i=0}^{N-1} a_{ik} T_k(t, m, \tau_V).
\end{equation}
$G_i(t, m, \tau_V)$ is the star-formation history of each galaxy $i$,
described as the stellar mass (in solar masses) formed $t$ years ago,
with metallicity $m$, and currently obscured by $\tau_V$ magnitudes in
the $V$ band. $T_k(t, m, \tau_V)$ is the star-formation history of one
of $N$ templates out of which we allow ourselves to build each
galaxy. We restrict to nonnegative combinations of these templates
($a_{ik} \ge 0$).

Each template is in turn expressed as a linear combination of a set of
instantaneous bursts:
\begin{equation}
T_k(t, m, d) = 
\sum_{l=0}^{N_t-1} 
\sum_{n=0}^{N_m-1} 
\sum_{p=0}^{N_\tau-1} 
t_{k,lmn} 
\delta^D(t-t_l)
\delta^D(m-m_n)
\delta^D(\tau_V-\tau_{V,p})
\end{equation}
That is, for each template $k$ a certain mass in stars $t_{k,lmn}$
(expressed in solar masses) were formed time $t_l$ ago with
metallicity $m_n$, and are at the time of observation obscured by
$\tau_V$ magnitudes of dust in the $V$ band.

As described in Section \ref{spectra}, these star-formation histories
can be converted into spectra using population synthesis models.

\subsection{Model for the ionized gas}

In the optical, the other important contribution to the flux other
than the stellar light is emission from ionized gas. Depending on how
it is ionized and what its metallicity is, the recombination of this
gas can yield rather different spectra.

Thus, we treat the gas as a sum of the same templates:
\begin{equation}
\label{emission}
E_i(m, q) = \sum_{i=0}^{N-1} a_{ik} S_k(m, q). 
\end{equation}
$E_i(m, q)$ is the amount of gas in each galaxy with metallicity $m$
and ionization parameter $q$.  $S_k(m, q)$ is the amount of such gas
in each of our $N$ templates.  We constrain the $a_{ik}$ in this
equation to be the same as that in Equation \ref{stellar} for the
star-formation histories. That is, each star formation history
templates $T_k{t, m, \tau_V}$ will be associated with the
corresponding distribution of ionized gas $S_k(m,q)$.

Again, each template is actually a sum of delta functions:
\begin{equation}
S_k(m, q) = 
\sum_{r=0}^{N_m-1} 
\sum_{s=0}^{N_q-1} 
t_{k,rs}'
\delta^D(m-m_r)
\delta^D(q-q_s)
\end{equation}
where for each metallicity $m_r$ and ionization parameter $q_s$ we 
have a prediction for the emergent line fluxes, which we describe in
Section \ref{spectra}.

\subsection{Predicted spectra from the star-formation histories and
ionized gas}
\label{spectra}

Note that we do not {\it have} to limit our spectral library to those
predicted by star-formation histories or the emission line models. In
fact, we can add arbitrary spectra in this step to account for
emission lines, polycyclic aromatic hydrocarbon (PAH) emission in the
mid-infrared, active galactic nuclei in the ultraviolet, or any other
constituent of galaxies.

@@

\subsection{Summary of the problem we want to solve}
\label{problem}

\section{Non-negative matrix factorization to find the best templates}
\label{nmf}

@@

%\include{fitting_tables}

\include{fitting_figures}
